---
title: "Home excericse - NOVA course on deep learning in remote sensing"
author: "Julius Wold"
format: gfm
tbl-cap-location: top
---

```{python}
#| include: FALSE

import os
from glob import glob
import itertools
import json

import pandas as pd
```

```{r}
#| include: FALSE

library(here)
library(tidyverse)
library(reticulate)
```

# Introduction

## Objective

Compare the effect of training a seedling detector on your own annotated dataset vs. the full dataset on the detector’s performance.

# Materials and methods

## Annotations

Two sets of annotations were used for model tranining: "My annotations" and "All annotations".

- "My annotations"
  - Annotations made by the author.
- "All annotations"
  - Annotations from all students merged.

```{python}
#| include: FALSE

def number_of_images(anno_data_path):
    n_train = len(glob(os.path.join(anno_data_path, "train", "images", "*.tif")))
    n_val = len(glob(os.path.join(anno_data_path, "val", "images", "*.tif")))
    n_images = {"train": n_train, "val": n_val, "all": n_train + n_val}

    return n_images


def number_of_boxes(anno_data_path):
    labels_train = glob(os.path.join(anno_data_path, "train", "labels", "*.txt"))
    n_train = sum(map(count_boxes, labels_train))
    labels_val = glob(os.path.join(anno_data_path, "val", "labels", "*.txt"))
    n_val = sum(map(count_boxes, labels_val))
    n_boxes = {"train": n_train, "val": n_val, "all": n_train + n_val}

    return n_boxes


def count_boxes(label):
    with open(label, "r") as f:
        n_boxes = len(f.readlines())

    return n_boxes


def anno_summary(anno_data_path):
    annotations_set = os.path.basename(anno_data_path).replace("_", " ").capitalize()
    n_images = number_of_images(anno_data_path)
    n_boxes = number_of_boxes(anno_data_path)
    df_index = pd.MultiIndex.from_product([["Number of images", "Number of trees"], ["Train", "Val", "Sum"]], names=["Count", "Split"])
    summary = pd.DataFrame({annotations_set: list(n_images.values()) + list(n_boxes.values())}, index=df_index)

    return summary
```

```{python}
#| include: FALSE

annotations = glob(os.path.join("..", "data", "annotated_data", "train", "*"))
summaries = pd.concat(map(anno_summary, annotations), axis=1)

```

```{r}
#| echo: FALSE
#| label: tbl-training
#| tbl-cap: Training data

knitr::kable(py$summaries)
```

## Model training

YOLOv8 models were trained using the dataset "My annotations" and the dataset "All annotations". A grid search were performed for model sizes "Nano", "Medium" and "Xtra large" and image sizes 256, 640 and 1024. The best mode for each dataset were selected using mAP@.5.

```{python}
#| include: FALSE

grid_search = pd.DataFrame(list(itertools.product(["yolov8n.pt", "yolov8m.pt", "yolov8x.pt"], [256, 640, 1024])), columns=["Model", "Image size"])
```

```{r}
#| echo: FALSE
#| label: tbl-search
#| tbl-cap: Grid search.

knitr::kable(py$grid_search)
```

## Model evaluation

The selected models from the model training were evaluated using machine-learning metris and domain metrics.

### ML metrics

Machine learning metrics were evaluated by testing the models against tiled test data.

## Domain metrics

$$
RMSE = \sqrt{\frac{\Sigma_{i=1}^{n}(y_i - \hat{y_i})^2}{n}}
$$

$$
MD = \frac{\Sigma_{i=1}^{n}(y_i - \hat{y_i})}{n}
$$

# Results & Discussion

## Model training

Models trained by the "My annotation" dataset achived better performance in training than the models trained on "All annotations". For "My annotation" the highest performance was acchived by the YOLOv8 nano model with image size 256 (mAP@.5 of 061 at epoch 133).

```{r}
my_perfomance <-
    here("models", "model_performance", "juliwold_home_exercise_sapling_detector_sapling_detector_table_data_my_annotations.csv") %>%
    read_csv(show_col_types = FALSE) %>%
    select(-data)
all_perfomance <-
    here("models", "model_performance", "juliwold_home_exercise_sapling_detector_sapling_detector_table_data_all_annotations.csv") %>%
    read_csv(show_col_types = FALSE) %>%
    select(-data)
```

```{r}
#| label: tbl-trainingperformance
#| tbl-cap: "Training performance of best model."
#| tbl-subcap: 
#|   - "My annotations"
#|   - "All annotations"
#| layout-ncol: 2

knitr::kable(my_perfomance)

knitr::kable(all_perfomance)
```

> Detailed view of trained models can be found in [this Comet project](<https://www.comet.com/juliwold/home-exercise-sapling-detector/view/uxR2erf0uJlERPXPjybwdN2yE/panels>).

```{r}
#| include: FALSE

collect_training_results <- function(model_path) {
    training_results <-
        here(model_path, "results.csv") %>%
        read_csv(show_col_types = FALSE) %>%
        mutate(model = basename(model_path), .before = 1)
}
```

```{r}
#| include: FALSE

training_results <-
    map(
        c("my_annotations", "all_annotations"),
        .f = ~{
            results_annotations <-
                here("models", .x) %>%
                list.files(full.names = TRUE) %>%
                map(collect_training_results) %>%
                bind_rows() %>%
                mutate(
                    annotation_set = .x, .before = 1,
                    model = str_remove(model, .x)
                )
        }
    ) %>%
    bind_rows()
```

```{r}
#| echo: FALSE
#| messages: FALSE
#| label: fig-training
#| fig-cap: Training results
#| width: 1024

training_plot <-
    training_results %>%
    mutate(
        model_size = str_extract(model, "yolov8[nmx]"),
        image_size = str_extract(model, "\\d{3,4}")
    ) %>%
    ggplot(aes(epoch, `metrics/mAP50(B)`, color = model_size, linetype = image_size)) +
    geom_line(alpha = 0.5) +
    geom_smooth(se = FALSE) +
    colorspace::scale_color_discrete_qualitative(palette = "Dark3") +
    # scale_colour_viridis_d() +
    facet_wrap(~annotation_set, ncol = 1, scales = "free")
training_plot
```

## Model evaluation

```{python}
#| include: FALSE
models = [
    "..\\models\\all_annotations\\all_annotations_yolov8m.pt_640",
    "..\\models\\my_annotations\\my_annotations_yolov8n.pt_256",
    "..\\models\\my_annotations\\my_annotations_yolov8n.pt_640",
]
```

### ML metrics

```{python}
#| include: FALSE

def collect_ml_metrics(model_path):
    with open(os.path.join(model_path, "val", "test_results.json"), "r") as f:
        ml_metrics = json.load(f)
    metrics_table = pd.DataFrame(ml_metrics, index=[os.path.basename(model_path)])

    return(metrics_table)
```

```{python}
#| include: FALSE

ml_metrics = pd.concat(map(collect_ml_metrics, models)).round(decimals=2)
```

```{r}
#| echo: FALSE
#| label: tbl-mlmetrics
#| tbl-cap: Machine learnining metrics.

knitr::kable(py$ml_metrics)
```

### Domain

```{python}
#| include: FALSE

def collect_domain_metrics(model_path):
    metrics = pd.read_csv(os.path.join(model_path, "predictions_processed", "metrics.csv"))
    metrics["Model"] = os.path.basename(model_path)
    return(metrics)
```

```{python}
#| include: FALSE

dm_all = pd.concat(map(collect_domain_metrics, models))
mi = pd.MultiIndex.from_frame(dm_all[["Model", "aoi_name"]])
```


```{python}
#| include: FALSE

domain_metrics = dm_all.set_index(mi).round()
domain_metrics = domain_metrics.drop(["aoi_name", "Model", "rmse_n", "rmse_n (%)","bias_n", "bias_n (%)",], axis = 1)
domain_metrics = domain_metrics.rename(columns={"rmse_dens": "RMSE", "rmse_dens (%)": "RMSE (%)", "bias_dens": "MD", "bias_dens (%)": "MD (%)"})
```

```{r}
#| echo: FALSE
#| label: tbl-comainmetrics
#| tbl-cap: Domain metrics.

knitr::kable(py$domain_metrics)
```

# Conclusion
